Report on a Safety Checker for Petri nets

### 2014-02-01 paper/workspace/main.tex
Added method Safety by Refinement with Lamport example

  # of tokens in place s           >= 0
  # of times transition t is fired >= 0

* Subprocedure \mathcal{C}:

@@

@@

The method Safety by Refinement applies trap conditions to check that a given Petri net \verb=N= never violates a property \verb=P=.
We present the method Invariant by example on a mutual exclusion algorithm by Delzanno2001.

\begin{verbatim}

* Code

bool m1 = false;
bool m2 = false;

// Thread 1:
while (true) {
  if (! m2 && ! test_and_set(m1)) {
    // critical section 1
    m1 = false
  }
  if (! m1 && ! test_and_set(m2)) {
    // critical section 2
    m2 = false
  }
}

// Thread 2:
while (true) {
  if (! m2 && ! test_and_set(m1)) {
    // critical section 1
    m1 = false
  }
  if (! m1 && ! test_and_set(m2)) {
    // critical section 2
    m2 = false
  }
}

Property: Thread 1 and Thread 2 are never in any critical section at the same time.

* Petri net:
\end{verbatim}
\begin{center}
\begin{tikzpicture}[node distance=2cm]
  \node[blue place, tokens=2, label=above:nc] (nc) {};
  \node[place, tokens=1, label=right:m2, below of=nc, xshift=+0.75cm] (s2) {};
  \node[place, tokens=1, label=left:m1, below of=nc, xshift=-0.75cm] (s1) {};
  \node[red place, label=above:cs1, left of=s1] (cs1) {};
  \node[red place, label=above:cs2, right of=s2] (cs2) {};
  \node[transition, label=above:t1] (t1) at ($(s1)!0.5!(cs1) + (0,1cm)$) {}
	edge [pre] (nc)
	edge [pre] (s1)
	edge [pre] (s2)
	edge [post] (cs1)
	edge [post] (s2);
  \node[transition, label=above:t2] (t2) at ($(s2)!0.5!(cs2) + (0,1cm)$) {}
	edge [pre] (nc)
	edge [pre] (s2)
	edge [pre] (s1)
	edge [post] (cs2)
	edge [post] (s1);
  \node[transition, label=below:t3] (t3) at ($(s1)!0.5!(cs1) - (0,1cm)$) {}
	edge [pre] (cs1)
	edge [post] (s1)
	edge [post,bend left,min distance=3cm, out=115, in=115] (nc);
  \node[transition, label=below:t4] (t4) at ($(s2)!0.5!(cs2) - (0,1cm)$) {}
	edge [pre] (cs2)
	edge [post] (s2)
	edge [post,bend right,min distance=3cm, out=-115, in=-115] (nc);
\end{tikzpicture}
\end{center}

\newpage
\begin{verbatim}
* Method Safety by Refinement:

\newpage
\begin{verbatim}
* Example
  - State constraints C:
    nc  =  2 - t1 - t2 + t3 + t4
    m1  =  1 - t1 + t3
    m2  =  1 - t2 + t4
    cs1 =  0 + t1 - t3
    cs2 =  0 + t2 - t4
    nc  >= 0
    m1  >= 0
    m2  >= 0
    cs1 >= 0
    cs2 >= 0
    t1  >= 0
    t2  >= 0
    t3  >= 0
    t4  >= 0

  - Assignment A:
    nc  = 0
    m1  = 0
    m2  = 0
    cs1 = 1
    cs2 = 1
    t1  = 1
    t2  = 1
    t3  = 0
    t4  = 0

    - Trap implications:

    cs1 => m1 \/ nc
    ^      ^     ^
    |      |     |
    |      |     nc \in S
    |      |
    |      m1 \in S
    |
    cs1 \in S

    cs2 => m2 \/ nc
    nc  => (m2 \/ cs1) /\ (m1 \/ cs2)
    m1  => (m2 \/ cs1) /\ (m1 \/ cs2)
    m2  => (m2 \/ cs1) /\ (m1 \/ cs2)

    - At least one of the intially marked places belongs to S:
    nc \/ m1 \/ m2

    - None of the marked places in A belongs to S:
    ~cs1 /\ ~cs2

  - Assignment A_\theta:
    nc  = false
    m1  = true
    m2  = true
    cs1 = false
    cs2 = false
\end{verbatim}

\newpage
\begin{verbatim}
  - Refinement constraint \delta:
    m1 + m2 >= 1
    ^    ^
    |    |
    ------
      |
      S = {m1, m2}, therefore \delta excludes assignment A in the next iteration

>>>>

  # of tokens in place s           >= 0
  # of times transition t is fired >= 0
\end{verbatim}

\begin{verbatim}
* Subprocedure \mathcal{C}:

@@

  Pseudocode:

  # Place equations
  for s in S:
    # initial number of tokens
    print(s + " = " + M0(s))
    for t in T:
      if (t, s) in E and not (s, t) in E:
        # input transition
        print(" + " + t)
      if (s, t) in E and not (t, s) in E:
        # output transition
        print(" - " + t)
    print("\n")

  # Non-negativity conditions
  for s in S:
    print(s + " >= 0\n")
  for t in T:
    print(t + " >= 0\n")
\end{verbatim}

The method Safety by Refinement applies trap conditions to check that a given Petri net \verb=N= never violates a property \verb=P=.

\begin{verbatim}
* Method Safety by Refinement:

\newpage
\begin{verbatim}
* Example
  - State constraints C and \neg P are the same as in the example for method Safety.

  - Assignment A:
    id1        = 0
    cs1        = 1
    idle1      = 0
    req1       = 0
    nid1       = 1
    id2        = 0
    cs2        = 1
    idle2      = 0
    req2       = 0
    after_you2 = 0
    await2     = 0
    u1         = 0
    u2         = 1
    u3         = 1
    v1         = 0
    v2         = 0
    v3         = 0
    v4         = 1
    v5         = 0
    v6         = 1

    - Trap implications:
                       v4              v3                 u2
                  ------------    --------------    --------------     
    id1        => (cs2 \/ id1) /\ (id1 \/ idle2) /\ (req1 \/ nid1)
    ^              ^      ^        ^      ^          ^       ^
    |              |      |        |      |          |       |
    |              |     id1 \in S |     idle2 \in S |      nid1 \in S
    |              |               |                 |
    |             cs2 \in S       id1 \in S         req1 \in S
    |
    id1 \in S

    cs1        => idle1 \/ id1
    idle1      => req1 \/ nid1
    req1       => cs1 \/ id2
    nid1       => (after_you2 \/ nid1) /\ (idle1 \/ id1)
    id2        => req2 /\ (cs1 \/ id2)
    cs2        => (idle2 \/ id2)
    idle2      => req2
    req2       => (cs2 \/ id1) /\ (cs2 \/ id1)
    after_you2 => id2 \/ await2
    await2     => id1 \/ idle2

    - At least one of the initially marked places belongs to S:
    id1 \/ id2 \/ idle1 \/ idle2

    - None of the marked places in A belongs to S:
    ~cs1 /\ ~cs2 /\ ~nid1
\end{verbatim}

\begin{verbatim}
  - Assignment A_\theta:
    id1        = true
    cs1        = false
    idle1      = false
    req1       = true
    nid1       = false
    id2        = true
    cs2        = false
    idle2      = false
    req2       = true
    after_you2 = true
    await2     = false
  - Refinement constraint \delta:
    id1 + id2 + req1 + req2 + after_you2 >= 1
    ^     ^     ^      ^      ^
    |     |     |      |      |
    ---------------------------
      |
      S = {iq1, iq2, req1, req2, after_you2}, therefore \delta excludes
      assignment A in the next iteration

  Pseudocode:

  # Trap implications
  for s in S:
    print(s + " => true")
    for t in s*:
      print(" /\ ( false")
      for p in t*:
        print(" \/ " + p)
     print(")")
    print("\n")

  # Initially marked
  print("false")
  for s in S:
    if M0(s) > 0:
      print(" \/ " + s)
  print("\n")

  # Unmarked in A
  for s in S:
    if A(s) = 0:
      print("~" + p + "\n")
\end{verbatim}

@@
